Applying State-free Reversible VAMPNets and Markov State Models to Learn Dynamics of DNA Oligonucleotides

### 2020-11-23 manuscript.tex
Updates from Overleaf

Following the analysis pipeline described above, we generate SRV-MSMs for each sequence. We identified seven kinetically relevant states that are captured within the resolution of the model. These include the fully hybridized state (H) in which all native base pairings are intact, the fully dissociated state (D), four "shifted" states (5S1, 5S2, 3S1, 3S2) where complementary base pairs form out-of-register, and a frayed state (F4), unique to GC-core, in which the four terminal A:T base pairs are unbound. Shifted states are only observed for AT-all and GC-end; state abbreviations indicate the direction of the shifted overhang (5' vs. 3') and the number of shifted motifs relative the native state (1 vs. 2). GC-mix displays two-state behavior within the resolution of the model, however we will show that the hybridization and dissociation are still characterized by an ensemble of shorter-lived states. Stationary probabilities in each of these states are shown in figure \ref{fig:allseq_table}. We also show that PCCA state assignments are capturing free energy minima in tICA space as independent states.

@@

The repetitive AT motifs in the AT-all and GC-end sequences produces a collection of out-of-register states similar to those shown in previous coarse-grained and all-atom studies \citep{Phys2014,  Romano2013DNADependence, Araque2016LatticeCooperativity, Xiao2019}. The thermodynamic stability of these states can be evaluated against experimental prediction by defining each structure in terms of "dangling ends" -- unpaired bases adjacent to the paired duplex -- and "inert tails" -- free bases that extend beyond the dangling end \citep{Michele2014EHybridization}. Dangling ends tend to have small stabilizing effects, and inert tails decrease stability as they increase in length -- 3' inert tails have stronger effects than 5' tails. Although sequence-dependent effects of inert tails are less well known, dangling end effects can be factored into nearest neighbor (NN) calculations \citep{Santalucia2004TM}. It is informative to compare our MSM predictions against NN calculations to understand how inert tail or kinetic effects cause deviation from these predictions. For AT-all, NN calculations (figure \ref{fig:NN_table}) predict that conformations in the 5' shifted states (5S1 and 5S2) are more energetically favorable than those in the 3' shifted state (3S1 and 3S2). We find qualitative agreement to our PCCA+ free energies, where inert tail effects likely contribute a 4-6.5 kJ/mol increase in free energy compared to dangling end predictions alone. For GC-end, we consider C:T and G:A mismatches in the GC-ends shifted states as non-interacting dangling ends such that each shifted conformation has four total dangling ends. Based on this treatment, NN calculations yield higher overall free energies (due to fewer native base pair contacts), but, contrary to PCCA+ results, predict that 3' out-of-register states to be more stable than 5' states. This leads us to believe that 5' vs. 3' differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weight NN stacking effects alone \citep{Doktycz1990ThermodynamicATGC}.

@@

Beyond state probabilities and free energy approximations, the coarse-grained MSM yields valuable discrete kinetic information in the form of transition probabilities between states. Figure \ref{fig:allseq_transition}!!!change ref!!! shows the probability of moving from one state to another within the MSM lag time. For AT-all, we observe approximately equal probability of transitioning from D to any other state, indicating that this is primarily a diffusion-driven process. Once a transition has been made, however, the 5' vs. 3' overhang and degree of shifting play an important role in determining whether the duplex will continue to shift out-of-register or re-dissociate. Transition probabilities are higher when moving towards a more aligned state than towards a more shifted state -- 5S1 $\rightarrow$ H is more favorable than 5S1 $\rightarrow$ 5S2, suggesting that these metastable shifted states play a more significant role in facilitating the hybridization process than dissociation. Furthermore, we see equal or higher transition probability from shifted states to the dissociated state (5S1 $\rightarrow$ D) than to more aligned states (5S1 $\rightarrow$ H), indicating that the shifting-hybridization process is frequently disrupted by complete dissociation. In particular, we observe that the transition probability 3S2 $\rightarrow$ D is 10x higher than 3S2 $\rightarrow$ 3S1. Indeed for GC-end, we observe no significant flux between the 5S2 state and the structural similar 5S1 state, indicating that the former acts more as a kinetic trap than a pathway to H. The 5S1 and 3S1 states still readily convert with H, however there is an order of magnitude lower probability of reaching these states from D. This indicates that inert tails inhibit out-of-register binding -- accounting in part for higher free energies discussed above -- but may not substantially disrupt out-of-register hybridization mechanisms along 5S1 $\rightarrow$ H and 3S1 $\rightarrow$ H once shifted states have formed.

@@

Taken together, our results indicate that AT-all hybridization and dissociation kinetics are substantially modulated by out-of-register pathways. A majority of hybridization events occur out-of-register, and, based on transition probabilities, 35\% of H $\rightarrow$ D pathways pass through some out-of-register state. While GC-end strands can access out-of-register states, only 11\% of native hybridization events pass through these states. These observations show that internal displacement mechanisms reported in previous simulation studies are capable of disrupting non-native pairing and correcting base pair mismatches without fully dissociating strands \citep{Romano2013DNADependence, Markegard2015, Maciejczyk2014DNAModel}. Experimentally, out-of-register states are suspected to occur, but are difficult to capture due to short lifetimes and subtle kinetic traces. To explicitly minimize out-of-register base pairing, similar AT repeat motif sequences have been padded by GCG clamps during experimental analysis \cite{Wyer2014KineticsAT-tracts}. Recent all-atom results identified similar out-of-register states as "deep kinetic traps" along the hybridization pathway for repetitive dGCGCGC hexamers \citep{Xiao2019}. Contrary to our kinetic results, however, "slithering" mechanisms along out-of-register tracts were not observed to be a dominant pathway, especially when compared to the high rates of slithering exhibited by the homogenous dGGGGGG strand. It is unclear whether these differences were a consequence of varied simulation conditions or whether GC repeat motifs are less susceptible to out-of-register transitions compared to AT motifs. This is possible given that stronger hydrogen binding in GC motifs may prevent fluctuation-driven rearrangement, however further computational and experimental studies are required to verify these differences.

@@

In examining spectroscopic T-jump signatures for these four sequences, we observed a significantly stretched fast response for AT-all, and a more subtle GC-end stretch relative to GC-core and GC-mix fast responses. For each sequence, the fast response is attributed to terminal base fraying, but stretching indicates a broader distribution of dynamics at this timescale. We expect that fraying in out-of-register conformations would occur at different rates than intact fraying. Furthermore, internal displacement mechanisms facilitating direct out-of-register transitions may produced similar responses contribute to the ensemble. Given that it is experimentally challenging to distinguish between these contributions, we cannot confirm whether these responses originate from some population of out-of-register configurations prior the temperature jump, sliding mechanisms during the dissociation process, or some combination of the two. In the context of our SRV-MSMs, we see qualitative agreement between more complex transition network and a more stretched response. We interrogate these differences by examining high temperature simulations later in this manuscript.

@@

\subsection{GC-core fraying} % not sure if I should make this its own section?

The GC-core sequence represents a departure from the dominant shifting dynamics observed for AT-all and GC-end. Instead, the dynamical analysis describes a hybridization/dissociation pathway facilitated by a unique, highly frayed state (F4). Previous studies suggests that once key contacts are made, the zippering mechanism ensures that the helix will quickly form outward \citep{Romano2013DNADependence, Yin2011KineticsHybridization}. Our results indicate, however, that the relative instability of AT bonds compared the the GC-core can interrupt this process and form a longer lived metastable state. This occurs during the dissociation process as well, where one half of the A:T base contacts are entirely broken for a substantial period of time before the full dissociation event occurs. We observe these events to occur with equal probability on either permutable side of the helix. The hybridized state has a 5x higher probability of transitioning into the frayed state within the lag time compared to transitions from the dissociated state. This is expected considering that this state is more accessible from an already bound helix. Moreover, once oligos are in F4, they are over 10x more likely to return to H than to D. Thus once a D $\leftrightarrow$ F4 transition has occurred, a F4 $\leftrightarrow$ H transition will likely proceed it. On the other hand, H $\leftrightarrow$ F4 events are more frequent but unlikely to initiate complete dissociation.

% try to split this up between an experimental section (all SRV-MSMs) and more detail on diffusion maps
\subsection{GC-core Comparison to Experiment}

% discuss computational comparisons needs work
Lattice model studies have shown that frayed intermediates make substantial contributions to the GC-core conformational ensemble \citep{Araque2016LatticeCooperativity, Phys2019}. Araque et al. defines a similar 8-mer sequence (dATGCGCAT) as non-two-state, where a stable, symmetrically A:T frayed state is a crucial part of the duplex transition path \citep{Araque2016LatticeCooperativity}. When examining all four sequences using T-jump IR and 2D IR spectroscopy, Sanstead et al. found that the GC-core had the highest deviation from two-state behavior during dissociation \citep{Sanstead2016}. As their lattice model did not consider previously mentioned shifted states, this intermediate state was defined by a high degree of fraying about the central core. While 1-2 base pair fraying was commonly observed for GC-mix and AT-all as well, lattice model predictions showed that GC-core had substantially more frayed base pairs \citep{Phys2019}. Variable T-jump measurements and Smoluchowski simulations on model 1D free energy landscapes showed that AT termini fraying was an effectively barrierless process characterized by rapid inter-conversion between all accessible frayed states \citep{Sanstead2018DirectDehybridization}. We see the same rapid fraying in simulation data -- which is too fast to be attributed to a converged SRV mode -- however we stipulate that this inter-conversion first relies on the formation of the the A:T bond nearest to the GC center.  Although this process occurs much slower than single A:T base bonding and breaking, it may be difficult to experimentally discern from the overall hybridization process which contains both G:C and A:T character and occurs on a similar timescale.

@@

\subsection{GC-mix displays more canonical hybridization and dissociation}
Although GC-mix dynamics are most similar to those of GC-core, we did not observe a converged slow mode corresponding to multi-base fraying behavior for GC-mix. Instead, we observed two modes converge, corresponding to the association/dissociation dynamic and diffusive behavior while strands are dissociated. As such, we designated this transitions as effectively two-state within the resolution of our model, however we do observe substantial fraying of the two AT termini in the simulation data. Although these frayed states may be too short-lived to resolve a distinct slow mode, this behavior shows qualitative agreement with experimental analysis of this sequence which attributed fraying prior to dissociation as a deviation from all-or-nothing behavior \citep{Sanstead2016}.  While AT-termini fraying is surely a prerequisite to dissociation, we find these states to be so common and fleeting that very few progress to full dissociation. On the other hand, the F4 state in GC-core had a substantial probability of fully dissociated rather reforming an intact helix. Furthermore, one or two base pair fraying does not fundamentally disrupt the helix in such a way that its re-formation is kinetically inhibited by the intermediate structures we present for GC-core.

Given the lack of a repetitive AT interior (as in AT-all and GC-end) or consecutive AT exterior (as in GC-core), we expect more canonical dynamics from GC-mix. For this analysis, we looked at qualitative trends in our trajectory data, paying close attention to the distances between matching WC-pairs (\ref{fig:GC-mix_transitions}). During hybridization, we observe the formations of some key base pair contacts before the full duplex forms. We observe that first contacts tend to involve one of the G:C bonds, which is not surprising that these are more evenly spaced out than in GC-core and GC-end. This behavior is indicative of a nucleation-zippering mechanism as has been reported in previous studies \citep{Wetmur1968KineticsDNA, Porschke1971CooperativeTransition, Yin2011KineticsHybridization}. We obverse dissociation events proceeded by two base pair fraying on one side of the duplex followed by more rapid dissociation of the central base pairs. There's evidence of a short-lived state composed of 2-4 base pair contacts immediately before full dissociation occurs. In contrast to the F4 state we observe in GC-core, these conformations do not form a distinct free-energy minima in SRV or tICA space, nor do they tend to reform intact duplexes. As a whole, these dynamics are similar to previously reported "fraying-peeling" mechanism \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding, Zgarbova2014BaseRNA}. We observed similar fast dynamics and transition states in the other three sequences, however they are more difficult to discern as they occur in concert with the longer lasting metastable states discussed above.

\begin{figure} %[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Figs/figs_imp/GC-mix_transitions.PNG}
        \caption{WC base pair distance and molecular renderings along two representative GC-mix hybridization and dissociation events.}
        \label{fig:GC-mix_transitions}
	\end{center}
\end{figure}

@@

% add new diffusion map section introduction
Although our SRV-MSMs provide useful insight into slow processes, we are also interested in the ensemble of configurations within PCCA+ macrostates. These configurations interchange significantly faster than the SRV and MSM lag times, but we can use a diffusion map embedding to glean a structural understanding of the macrostate population. Diffusion maps generate a low dimensional embedding of the data into high variance structural modes and are well-suited to find subtle differences in temporally disconnected data \citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps}. We were particularly interested in using this method to explore the 5s1 and F4 macrostates as these represent the most popluated and kinetically relevant macrostates for AT-all, GC-end, and GC-core.

% use this to substantiate the shifted fast response
\subsubsection{5s1 state analysis}
While examining molecular renderings, we noticed that a significant proportion of GC-end 5s1 configurations retained one native G:C bond, even when all available A:T bonds were formed out-of-register. We would not necessarily expect duplexes to sacrifice helical conformational entropy in order to facilitate termini bonding. To compare how these state populations differ between GC-end and AT-all, we employed diffusion maps built on an equal sampling of 5000 conformations from the 5S1 state of both sequences. We used all 100 intermolecular distances (as opposed to the 55 permutation free coordinates used to construct SRV-MSM) as our distance metric, making it easier to discern structures that form on either permutable end of the shifted conformation. This created a degenerate 2nd and 3rd diffusion modes, with nearly equal eigenvalues, differentiating between looping at the identical "top" and "bottom" of the strands. In Figure \ref{fig:GC-end_dmaps} we present the first two non-trivial diffusion map eigenfunctions and show representations of the degenerate third coordinate in the SI (Figure \ref{fig:GC-end_dmaps_full}). Diffusion maps built from samples of the 3' shifted states are also shown in the SI (Figure \ref{fig:GC-end_dmaps_3prime}).

The first diffusion mode clearly delineates between the GC-end and AT-all shifted conformations and correlates highly with the average distance between the 3' end and its shifted complementary pair. This reveals that the mismatch C:T-pairs are never bound -- a consequence of the 3spn2 excluded volume interaction -- whereas the AT-all pairs are mostly bound with occasional fraying indicated by small AT-all overlap in the GC-end region. This effect may be exaggerated given that C and T base pairs are assigned slightly higher excluded volume radii in 3spn2 \citep{Hinckley2013AnHybridization}. The second diffusion mode, which correlates highly with the average distance between 3' and 5' ends, has higher values for GC-end than AT-all. Because the GC-end termini do not bind out of register, we find that they are readily able to form stabilizing contacts despite the shifted conformation of the duplex as a whole. These "shifted-loop" bonds are shown to be uniquely stable for GC-end conformations in the 5' shifted state, and their existence in the simulations confirmed by molecular renderings of these regions Although AT-all shifted ends tend to stay bound out-of-register, the second diffusion coordinate shows some population of inert tails that fold back onto the helix.

% need to work on this if adding to this section (might belong somewhere else or not at all?
Although internal base pair mismatches can cause substantial conformational distortions such as kinking, terminal mismatches have been shown to be slightly stabilizing and have a minimal effect on helical character \citep{Santalucia2004TM, DiMichele2014EffectHybridization}. In the context of the 3spn2 model, these ends are accounted for via intra-strand base stacking and inter-strand cross-stacking interactions \citep{Hinckley2013AnHybridization}. The only direct interaction between non Watson-Crick (WC) basepairs is parameterized by isotropic excluded volume potential, which is likely more simplistic than the true mismatch interaction.

@@

@@

We applied a similar approach to investigate the structural composition of the F4 macrostate. We build diffusion maps using 10000 frames sampled from the macrostate \ref{fig:GC-core_dmaps}. This time, we set our distance metric to the same permutation-free coordinates we used to build the SRV-MSMs. Again, we were able to identify a combination of physical coordinates that closely correlated to the first two non-trivial diffusion modes. We found that a larger distance between the third and fourth A:T pairs increased the PCCA+ probability of inclusion into macrostate. This distance also correlated closely with the second diffusion map mode. Interestingly, we observed that the first non-trivial mode -- the feature that describes the most structural diversity in the system -- corresponds to difference in "overlap" distance between adjacent A or T bases and the GC core. In these conformations, one of the strands maintains some helical character while the other twist out of place, resulting in WC bonds being obstructed by the oligo backbone. These states represent another potential way in which the hybridization process (or helix reformation) can be kinetically frustrated. We observed this mode to be mostly symmetric, however there is slight tendency for the 3'T end to fray farther out of place relative to its 5'A counterpart. This might be another consequence of differential excluded volume radii in the 3spn2 force field.

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Figs/figs_imp/GC-core_dmaps.png}
        \caption{Plotting the first two non-trivial diffusion map eigenfunctions. Color maps show the probability that conformations are clustered into the frayed macrostate, maximum distance between 4th AT basepairs, and the the overlap between adjacent AT basepairs and the GC core.}
        \label{fig:GC-core_dmaps}
	\end{center}
\end{figure}

@@

Given differences in temperature and ensemble distributions between Tmelt simulations and T-jump experiments, we found it difficult to make direct timescales comparisons based on our equilibrium models alone. To supplement our analysis, we ran short simulations initialized in the hybridized state at a series of elevated temperatures for each sequence. We derived relaxation times for a "slow" dissociation response and "fast" fraying response at each temperature and compared these with experimental temperature-dependent relaxation fits for GC-end and GC-core. Additionally, we repeated the protocol in Sandstead et al. to generate temperature series data for AT-all and GC-mix sequences \citep{Sanstead2018DirectDehybridization}.

@@

>>>>

Following the analysis pipeline described above, we generate SRV-MSMs for each sequence. We identified seven kinetically relevant states that are captured within the resolution of the model. These include the fully hybridized state (H) in which all native base pairings are intact, the fully dissociated state (D), four "shifted" states (5S1, 5S2, 3S1, 3S2) where complementary base pairs form out-of-register, and a frayed state (F4), unique to GC-core, in which the four terminal A:T base pairs are unbound. Shifted states are only observed for AT-all and GC-end; state abbreviations indicate the direction of the shifted overhang (5' vs. 3') and the number of shifted motifs relative the native state (1 vs. 2). GC-mix displays two-state behavior within the resolution of the model, however we will show that the hybridization and dissociation are still characterized by an ensemble of shorter-lived states. Stationary probabilities in each of these states are shown in figure \ref{fig:allseq_table}. We also show that PCCA+ clustering assigns states to relative free energy minima in the tICA space.

@@

The repetitive AT motifs in the AT-all and GC-end sequences produces a collection of out-of-register states similar to those shown in previous coarse-grained and all-atom studies \citep{Phys2014,  Romano2013DNADependence, Araque2016LatticeCooperativity, Xiao2019}. The thermodynamic stability of these states can be evaluated against experimental prediction by defining each structure in terms of "dangling ends" -- unpaired bases adjacent to the paired duplex -- and "inert tails" -- free bases that extend beyond the dangling end \citep{Michele2014EHybridization}. Dangling ends tend to have small stabilizing effects, and inert tails decrease stability as they increase in length -- 3' inert tails have stronger effects than 5' tails. Although sequence-dependent effects of inert tails are less well known, dangling end effects can be factored into nearest neighbor (NN) calculations \citep{Santalucia2004TM}. It is informative to compare our MSM predictions against NN calculations to understand how inert tail and kinetic effects may cause deviation from these predictions. For AT-all, NN calculations (figure \ref{fig:NN_table}) predict that conformations in the 5' shifted states (5S1 and 5S2) are more energetically favorable than those in the 3' shifted state (3S1 and 3S2). We find qualitative agreement to our PCCA+ free energies, where inert tail effects likely contribute a 4-6.5 kJ/mol increase in free energy compared to dangling end predictions alone. For GC-end, we consider C:T and G:A mismatches in the GC-ends shifted states as non-interacting dangling ends such that each shifted conformation has four total dangling ends. Based on this treatment, NN calculations yield higher overall free energies due to fewer native base pair contacts.  Contrary to PCCA+ results, however, 3' out-of-register states to be more stable than 5' states. This leads us to believe that 5' vs. 3' differences -- attributed to some combination of 5' tails preferentially stacking on the core duplex and 3' tails perturbing the duplex structure -- may out-weight NN stacking effects alone \citep{Doktycz1990ThermodynamicATGC}.

@@

In addition to state probabilities and free energy approximations, the coarse-grained MSM yields valuable discrete kinetic information in the form of transition probabilities between states. Figure \ref{fig:allseq_transition} shows the probability of moving from one state to another within the MSM lag time. For AT-all, we observe approximately equal probability of transitioning from D to any other state, indicating that this is primarily a diffusion-driven process. Once a transition has been made, however, the 5' vs. 3' overhang and degree of shifting play an important role in determining whether the duplex will continue to shift out-of-register or re-dissociate. Transition probabilities are higher when moving towards a more aligned state than towards a more shifted state -- 5S1 $\rightarrow$ H is more favorable than 5S1 $\rightarrow$ 5S2 -- suggesting that these metastable shifted states play a more significant role in facilitating the hybridization process than dissociation. Furthermore, we see equal or higher transition probability from shifted states to the dissociated state (5S1 $\rightarrow$ D) than to more aligned states (5S1 $\rightarrow$ H), indicating that the shifting-hybridization process is frequently disrupted by complete dissociation. In particular, we observe that the transition probability 3S2 $\rightarrow$ D is 10x higher than the 3S2 $\rightarrow$ 3S1 path to native hybridization. Indeed, for GC-end we observe no significant flux between the 5S2 state and the structural similar 5S1 state, indicating that the former acts more as a kinetic trap than a pathway to H. The 5S1 and 3S1 states still readily convert with H, however there is an order of magnitude lower probability of reaching these states from D. This indicates that inert tails inhibit out-of-register binding -- accounting in part for higher free energies discussed above -- but may not substantially disrupt out-of-register hybridization mechanisms along 5S1 $\rightarrow$ H and 3S1 $\rightarrow$ H once shifted states have formed.

@@

Taken together, our results indicate that AT-all hybridization and dissociation kinetics are substantially modulated by out-of-register pathways. A majority of hybridization events occur out-of-register, and, based on transition probabilities, 35\% of D $\rightarrow$ H pathways pass through some out-of-register state. While GC-end strands can access out-of-register states, only 11\% of native hybridization events pass through these states. These observations show that internal displacement mechanisms reported in previous simulation studies are capable of disrupting non-native pairing and correcting base pair mismatches without fully dissociating strands \citep{Romano2013DNADependence, Markegard2015, Maciejczyk2014DNAModel}. Experimentally, out-of-register states are suspected to occur, but are difficult to capture due to short lifetimes and subtle kinetic traces. To explicitly minimize out-of-register base pairing, similar AT repeat motif sequences have been padded by GCG clamps during experimental analysis \cite{Wyer2014KineticsAT-tracts}. Recent all-atom results identified similar out-of-register states as "deep kinetic traps" along the hybridization pathway for repetitive dGCGCGC hexamers \citep{Xiao2019}. Contrary to our kinetic results, however, "slithering" mechanisms along out-of-register tracts were not observed to be a dominant pathway, especially when compared to the high rates of slithering exhibited by the homogenous dGGGGGG strand. It is unclear whether these differences were a consequence of varied simulation conditions or whether GC repeat motifs are less susceptible to out-of-register transitions compared to AT motifs. This is possible given that stronger hydrogen binding in GC motifs may prevent fluctuation-driven rearrangement, however further computational and experimental studies are required to verify these differences.

@@

In examining spectroscopic T-jump signatures for these four sequences, we observed a significantly stretched fast response for AT-all, and a more subtle GC-end stretch relative to GC-core and GC-mix fast responses. For each sequence, the fast response is attributed to terminal base fraying, but stretching indicates a broader distribution of dynamics at this timescale. We expect that fraying would still occur in out-of-register conformations but at a different rate than intact fraying. Furthermore, internal displacement mechanisms facilitating direct out-of-register transitions may produced similar responses contributing to the ensemble. Given that it is experimentally challenging to distinguish between these contributions, we cannot confirm whether these responses originate from some population of out-of-register configurations prior the temperature jump, shifting mechanisms during the dissociation process, or some combination of the two. In the context of our SRV-MSMs, we see qualitative agreement between more complex transition network and a more stretched response. We interrogate these differences by examining high temperature simulations later in this manuscript. (Can discuss more details about fig \ref{fig:fast_stretching} here and potentially include this fig in the main text)?

@@

\subsection{GC-core metastability} % not sure if I should make this its own section?

The GC-core sequence represents a departure from the dominant shifting dynamics observed for AT-all and GC-end. Instead, the dynamical analysis describes a hybridization/dissociation pathway facilitated by a unique, highly frayed state (F4). Previous studies suggests that once key contacts are made, the zippering mechanism ensures that the helix will quickly form outward \citep{Romano2013DNADependence, Yin2011KineticsHybridization}. Our results indicate, however, that the relative instability of A:T bonds compared the GC core can interrupt this process and form a longer lived metastable state. This occurs during the dissociation process as well, where one half of the A:T base contacts are entirely broken for a substantial period of time before the full dissociation event occurs. We observe these events to occur with equal probability on either permutable end of the helix. The F4 state is more accessible from an already bound helix, and once oligos are in F4, they are over 20x more likely to return to H than to D. Thus once a D $\leftrightarrow$ F4 transition has occurred, a F4 $\leftrightarrow$ H transition will likely proceed it. On the other hand, H $\leftrightarrow$ F4 events are more frequent but unlikely to initiate complete dissociation.

% discuss computational comparisons needs work
Lattice model studies have shown that frayed intermediates make substantial contributions to the GC-core conformational ensemble \citep{Araque2016LatticeCooperativity, Phys2019}. Araque et al. defines a similar 8-mer sequence (dATGCGCAT) as non-two-state, where a stable, symmetrically A:T frayed state is a crucial part of the duplex transition path \citep{Araque2016LatticeCooperativity}. When examining all four sequences using T-jump IR and 2D IR spectroscopy, Sanstead et al. found that the GC-core had the highest deviation from two-state behavior during dissociation \citep{Sanstead2016}. As their lattice model did not consider previously mentioned shifted states, this intermediate state was defined by a high degree of fraying about the central core. While 1-2 base pair fraying was commonly observed for GC-mix and AT-all as well, lattice model predictions showed that GC-core had substantially more frayed base pairs \citep{Phys2019}. Variable T-jump measurements and Smoluchowski simulations on model 1D free energy landscapes showed that AT termini fraying was an effectively barrierless process characterized by rapid inter-conversion between all accessible frayed states \citep{Sanstead2018DirectDehybridization}. We see the same rapid fraying in simulation data -- which is too fast to be attributed to a converged SRV mode -- however we stipulate that this inter-conversion first relies on the slower formation of the the A:T bond nearest to the GC center.  Although this process occurs much slower than single A:T base bonding and breaking, it may be difficult to experimentally discern from the overall hybridization process which contains both G:C and A:T character and occurs on a similar timescale.

@@

\subsection{GC-mix displays more canonical hybridization and dissociation}
Although GC-mix dynamics are most similar to those of GC-core, we did not observe a converged slow mode corresponding to multi-base fraying behavior for GC-mix. Instead, we observed two modes converge, corresponding to the association/dissociation dynamic and diffusive behavior while strands are dissociated, respectively. Given that this second mode did not inform the hybridization process, we designated this transitions as effectively two-state within the resolution of our model. We did, however, we do observe substantial fraying of the two AT termini in the simulation data. Although these frayed states may be too short-lived to resolve a distinct slow mode, this behavior shows qualitative agreement with experimental analysis of this sequence which attributed fraying prior to dissociation as a deviation from all-or-nothing behavior \citep{Sanstead2016}.  While AT-termini fraying is surely a prerequisite to dissociation, we find these states to be so common and fleeting that very few progress to full dissociation compared to the F4 state. Furthermore, one or two base pair fraying does not fundamentally disrupt the helix in such a way that its re-formation is kinetically inhibited by the intermediate structures we present for GC-core.

Given the lack of a repetitive AT interior (as in AT-all and GC-end) or consecutive AT exterior (as in GC-core), we expect more canonical dynamics from GC-mix. For this analysis, we looked at qualitative trends in our trajectory data, paying close attention to the distances between matching WC-pairs (\ref{fig:GC-mix_transitions}). During hybridization, we observed the formations of some key base pair contacts before the full duplex formed. Specifically, first contacts tended to involve one of the G:C bonds and at least one neighboring A:T. This behavior is indicative of a nucleation-zippering mechanism as has been reported in previous studies \citep{Wetmur1968KineticsDNA, Porschke1971CooperativeTransition, Yin2011KineticsHybridization}. For dissociation events, we noted two base pair fraying on one or both sides of the duplex followed by more rapid dissociation of the central base pairs. There's evidence of a short-lived state composed of 2-4 base pair contacts immediately before full dissociation occurs. In contrast to the F4 state we observe in GC-core, these conformations do not form a distinct free-energy minima in SRV or tICA space, nor do they tend to reform intact duplexes. As a whole, these dynamics are similar to previously reported "fraying-peeling" mechanism \citep{Wong2008TheSimulations, Perez2010Real-timeUnfolding, Zgarbova2014BaseRNA}. We observed similar fast dynamics and transition states in the other three sequences, however they are more difficult to discern as they occur in concert with the longer lasting metastable states discussed above.

\begin{figure} %[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Figs/figs_imp/GC-mix_transitions.PNG}
        \caption{WC base pair distance and molecular renderings along two representative GC-mix hybridization and dissociation events. Permutable distances are reflected across the x-axis and show that fraying in the hybridized states tends to occur independently on either end of the duplex until fully dissociation occurs. The SRV-MSM state is shown along the x-axis, indicating the point at which a transition has been determined. Fluctuation between states is common during the transition.}
        \label{fig:GC-mix_transitions}
	\end{center}
\end{figure}

@@

% add new diffusion map section introduction
Although our SRV-MSMs provide useful insight into slow processes, we were also interested in the ensemble of configurations within PCCA+ macrostates. These configurations interchange significantly faster than the SRV and MSM lag times, but we can use a diffusion map embedding to glean a structural understanding of the macrostate population. Diffusion maps generate a low dimensional embedding of the data into high variance structural modes and are well-suited to find subtle differences in temporally disconnected data \citep{Coifman2006DiffusionMaps, Ferguson2010SystematicMaps}. We were particularly interested in using this method to explore the 5s1 and F4 macrostates as these represent the most popluated and kinetically relevant macrostates for AT-all, GC-end, and GC-core.

% use this to substantiate the shifted fast response
\subsubsection{5s1 state analysis}
While examining molecular renderings, we noticed that a significant proportion of GC-end 5s1 configurations retained one native G:C bond, even when all available A:T bonds were formed out-of-register. We would not necessarily expect duplexes to sacrifice helical conformational entropy in order to facilitate termini bonding. To compare how these state populations differ between GC-end and AT-all, we employed diffusion maps built on an equal sampling of 5000 conformations from the 5S1 state of both sequences. We used all 100 intermolecular distances (as opposed to the 55 permutation free coordinates used to construct SRV-MSM) as our distance metric, making it easier to discern structures that form on either permutable end of the shifted conformation. This created degenerate 2nd and 3rd diffusion modes, with nearly equal eigenvalues, differentiating effects at the identical "top" and "bottom" of the strands. In Figure \ref{fig:GC-end_dmaps} we present the first two non-trivial diffusion map eigenfunctions and show representations of the degenerate third coordinate in the SI (Figure \ref{fig:GC-end_dmaps_full}). Diffusion maps built from samples of the 3S1 states are also shown in the SI (Figure \ref{fig:GC-end_dmaps_3prime}).

The first diffusion mode clearly delineates between the GC-end and AT-all shifted conformations and correlates highly with the average distance between the 3' end and its shifted complementary pair. This reveals that the mismatched C:T pairs are never bound -- a consequence of the 3spn2 excluded volume interaction -- whereas the AT-all pairs are mostly bound with occasional fraying indicated by small AT-all overlap in the GC-end region. This effect may be exaggerated given that C and T base pairs are assigned slightly higher excluded volume radii in 3spn2 \citep{Hinckley2013AnHybridization}. The second diffusion mode, which correlates highly with the average distance between 3' and 5' ends, has higher values for GC-end than AT-all. Because the GC-end termini do not bind out of register, we find that they are readily able to form stabilizing contacts despite the shifted conformation of the duplex as a whole. These "shifted-loop" bonds are shown to be uniquely stable for GC-end conformations in the 5' shifted state, and their existence in the simulations is confirmed by molecular renderings sampled from these regions. Although AT-all shifted ends tend to stay bound out-of-register, the second diffusion coordinate shows some population of inert tails that fold back onto the helix.

% need to work on this if adding to this section (might belong somewhere else or not at all?
%Although internal base pair mismatches can cause substantial conformational distortions such as kinking, terminal mismatches have been shown to be slightly stabilizing and have a minimal effect on helical character \citep{Santalucia2004TM, DiMichele2014EffectHybridization}. In the context of the 3spn2 model, these ends are accounted for via intra-strand base stacking and inter-strand cross-stacking interactions \citep{Hinckley2013AnHybridization}. The only direct interaction between non Watson-Crick (WC) basepairs is parameterized by isotropic excluded volume potential, which is likely more simplistic than the true mismatch interaction.

@@

\begin{figure}[ht!]
	\begin{center}
        \includegraphics[width=\textwidth]{Figs/figs_imp/GC-core_dmaps.png}
        \caption{Plotting the first two non-trivial diffusion map eigenfunctions. Color maps show the probability that conformations are clustered into the frayed macrostate, maximum distance between 4th AT basepairs, and the overlap between adjacent A:T basepairs and the GC core.}
        \label{fig:GC-core_dmaps}
	\end{center}
\end{figure}

@@

We applied a similar approach to investigate the structural composition of the F4 macrostate. We build diffusion maps using 10000 frames sampled from the macrostate (figure \ref{fig:GC-core_dmaps}). This time, we set our distance metric to the same permutation-free coordinates we used to build the SRV-MSMs. Again, we were able to identify a combination of physical coordinates that closely correlated to the first two non-trivial diffusion modes. We found that a larger distance between internal A:T pairs increased the PCCA+ probability of inclusion into macrostate. This distance also correlated closely with the second diffusion map mode. Interestingly, we observed that the first non-trivial mode -- the feature that describes the most structural diversity in the system -- corresponds to difference in "overlap" distance between adjacent A or T bases and the GC core. We defined overlap as the distance between a G/C and the A/T adjacent to its native complement. In these conformations, one of the strands maintains some helical character while the other twists out of place, resulting in WC bonds being obstructed by the oligo backbone. These states represent another potential way in which the hybridization process (or helix reformation) can be kinetically frustrated. We observed this mode to be mostly symmetric, however there is slight tendency for the 3'T end to fray farther out of place relative to its 5'A counterpart. This might be another consequence of differential excluded volume radii in the 3spn2 force field.

@@

Given differences in temperature and ensemble distributions between Tmelt simulations and T-jump experiments, we found it difficult to make direct timescales comparisons based on our equilibrium models alone. To supplement our analysis, we ran short simulations initialized in the hybridized state at a series of elevated temperatures for each sequence. We derived relaxation times for a "slow" dissociation response and "fast" fraying response at each temperature and compared these with experimental temperature-dependent relaxation fits for GC-end and GC-core. Additionally, we repeated the protocol in \citet{Sanstead2018DirectDehybridization} to generate temperature series data for AT-all and GC-mix sequences.

@@

\subsection{VAMP-2 Scoring}
Optimization of hyperparameters including feature choice, number of microstates, and number of slow modes was evaluated with via the VAMP-2 score. In the equations below, we show how covariances are obtained from some featurization $\chi$ of a time series $x_t$ and its time-lagged pairs $x_{t+\tau}$. The VAMP-2 score can then be found for $\chi$ by applying the VAMP principle with cross-validation.
